---
title: "Lecture 5 - Data Manipulation"
output: html_document
---

```{r}
library(tidyverse)

options(tibble.print_min = 3)

getwd()
```

# Import and organize data 
```{r}
litters_data = read_csv(file = "./sample data/FAS_litters.csv",
  col_types = "ccddiiii") #c indicates the first col is a character, d = decimal/double, i = integer
  
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./sample data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)
```

# Select & Rename columns 

```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive) # select by writing the dataset name followed by a list of col names to keep 

select(litters_data, group:gd_of_birth) #instead of listing as we first did, you can identify a range of columns to keep using :

select(litters_data, -pups_survive) #instead of indicating which to keep, can indicate which to remove by adding a neg sign in front of each listed col name

select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number) #selects & renames at the same time. new name = old name

rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number) #rename variables if don't need to also select certain cols. this is more useful for changing only a couple vars bc by hand

select(litters_data, litter_number, group, everything()) #currently the group values are on the far left hand side of the dataset. use this to select and move all of the litter_number obs from the second-to-left column position to the far left position. 

select (litters_data, starts_with("gd")) #select all obs in the litters_data that starts with "gd". Other commands include "contains(), ends_with(), starts_with()" - can look up more in the help tab 
```

#Learning Assessment
```{r}
select(pups_data, litter_number, sex, pd_ears) #select the columns containing litter number, sex, and PD ears
```

Note: using 'select' will export a dataframe even if you only select one column. Mostly this is fine, but sometimes you want the vector stored in the column. To pull a single variable, use 'pull'.

#Filter

```{r}
#examples of filters that select obs if they're >,==, etc a certain value
filter(pups_data, gd_of_birth == 20)        
filter(pups_data, pups_born_alive >= 2)
filter(pups_data, pups_survive != 4)          # ! negates what follows it 
filter(pups_data, !(pups_survive == 4))
filter(pups_data, group %in% c("Con7", "Con8"))    #could also write group == "Con7"" | group == "Con8"   |=or, &=and 
filter(pups_data, group == "Con7" & gd_of_birth == 20)
filter(pups_data, !is.na(wt_increase))        # na = missing obs
```

# learning Assessment

```{r}
filter(pups_data, sex == 1)   #Filter to include only pups with sex 1
filter(pups_data, sex == 2, pd_walk < 11) #include only pups with PD walk less than 11 and sex equal to 2
```

# Mutate
Use 'mutate' if want to change, select, or create new columns
```{r}
# the following doesn't necessarily save the additions to the file. To do that you'd add "newdatasetname = ..." in front of the mutate command
mutate(litters_data, 
  wt_gain = gd18_weight - gd0_weight, #create new var 'wt_gain' that is a function of existing vars
  group = tolower(group)
)
```

# Learning Assessment

```{r}
mutate(pups_data, pivot_minus7 = pd_pivot - 7) #Create new var that subtracts 7 from PD pivot
mutate(pups_data, pd_sum = pd_ears + pd_eyes + pd_pivot + pd_walk) #Create new var that is the sum of all the PD variables
```

#Arrange

```{r}
#reorder columns by listing the var names 
head(arrange(litters_data, group, pups_born_alive), 10)
```

There are a couple of options for doing multi-step data manipulation (importing, cleaning, selecting, etc):
*define intermediate datasets (or overwrite data at each stage)
*nest function calls

#PIPE %>% = to do a multi-step procedure in a simpler way 

we want to load the data, clean the column names, remove 'pups_survive', and create  new var 'wt_gain'

##Option 1 = without using mutate to create an edited dataset:
```{r}
litters_data_raw = read_csv("./sample data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars = mutate(litters_data_selected_cols, 
  wt_gain = gd18_weight - gd0_weight,
  group = tolower(group))
litters_data_with_vars
```

##Option 2 = using mutate to create an edited dataset 'litters_data_clean' but still a little cumbersome bc have to read commands from inside out:
```{r}
litters_data_clean = 
  mutate(
    select(
      janitor::clean_names(
        read_csv("./sample data/FAS_litters.csv", col_types = "ccddiiii")
        ), 
    -pups_survive
    ),
  wt_gain = gd18_weight - gd0_weight,
  group = tolower(group)
  )
litters_data_clean
```

##Option 3: using to mutate in simplest fashion (read outside in) to create an edited dataset 'litters_data' by selecting everything except pups_survive and then creating wt_gain and use %>% to plug it into the following command 
```{r}
litters_data = 
  read_csv("./sample data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>% 
#the following lines take the output from the previous three lines and plugs it into later commands
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = tolower(group))
litters_data


#the folowing does the same thing as above but uses a placeholder dataset '.' which is created by 'dat = .' :
litters_data = 
  read_csv("./sample data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names(dat = .) %>%
  select(.data = ., -pups_survive) %>%
  mutate(.data = .,
    wt_gain = gd18_weight - gd0_weight,
    group = tolower(group))
#the placeholder is more necessary/useful in the following example:
litters_data %>%
  lm(wt_gain ~ pups_born_alive, data = .) %>%
  broom::tidy()
```

# Learning Assessment:
```{r}
pups_data = 
  read_csv("./sample data/FAS_pups.csv", col_types = "ciiiii") %>%
  janitor::clean_names() %>%
  filter(sex == 1) %>%
  select(-pd_ears) %>%
  mutate(pd_pivot_gt7 = pd_pivot > 7)

```
